Accelerating structure-function mapping using the ViVa webtool to mine natural variation

### 2019-04-21 manuscript/04-methods.Rmd
submitted revisions

### Germplasm Accession Information
A dataset of each of the 1135 accessions including CS stock numbers and geographic location where the samples were collected was retrieved from the 1001 Genomes website at http://1001genomes.org/accessions.html, via the download link at the bottom of the page. This data file has been embedded in the R package as `accessions`.

@@

(ref:SNPeff-order) **Rank order of the strength of functional effects** Effect classes were ordered by subjective prediction of average strength of effect on gene function. Strength was then assigned to each effect on an integer scale.
```{r SNPeff-order, fig.cap="(ref:SNPeff-order)"}
SNPeff_order$effect <- factor(as.character(SNPeff_order$effect), levels = as.character(SNPeff_order$effect), ordered = TRUE)

@@

## Testing
A group consisting of undergraduate and graduate students, postdoctoral and postbaccalaureate researchers (the authors of this publication) were assembled to test the functionality of ViVa. Testers were asked to use the ViVa web interface to analyze the genetic variation within a gene family or clade within a family of the auxin nuclear signaling pathway. Testers were provided with a brief vignette on how to use ViVa to formulate new hypotheses or support existing hypotheses about gene function and evolution, which has been expanded to the "An Overview of ViVa" section. The testers had background knowledge of these genes as members of laboratories studying auxin nuclear signaling. Testers met weekly with the developers to discuss their user experience and their findings. Tester experiences, issues, and suggestions were incorporated into the ViVa software. The results collected by these analyses are summarized in the "Visualizing Variation within the auxin signaling pathway" section and [Supplemental Data](#si).

>>>>

### Germplasm Accession Information
A dataset of each of the 1135 accessions including CS stock numbers and geographic locations where the samples were collected was retrieved from the 1001 Genomes website at http://1001genomes.org/accessions.html, via the download link at the bottom of the page. This data file has been embedded in the R package as `accessions`.

@@

(ref:SNPeff-order) **Rank order of the strength of functional effects** Variant effect classes were ordered by subjective prediction of average strength of effect on gene function. Strength was then assigned to each effect on an integer scale.
```{r SNPeff-order, fig.cap="(ref:SNPeff-order)"}
SNPeff_order$effect <- factor(as.character(SNPeff_order$effect), levels = as.character(SNPeff_order$effect), ordered = TRUE)

@@

## Testing
A group consisting of undergraduate and graduate students, post-doctoral and post-baccalaureate researchers (the authors of this publication) were assembled to test the functionality of ViVa. Testers were asked to use the ViVa web interface to analyze the genetic variation at the clade or whole gene family level of the auxin nuclear signaling pathway. Testers were provided with a brief vignette on how to use ViVa to formulate new hypotheses or support existing hypotheses about gene function and evolution, which has been expanded to the "An Overview of ViVa" section. The testers had background knowledge of these genes as members of laboratories studying auxin nuclear signaling. Testers met weekly with the developers to discuss their user experience and their findings. Tester experiences, issues, and suggestions were incorporated into the ViVa software. The results collected by these analyses are summarized in the "Visualizing Variation within the auxin signaling pathway" section and [Supplemental Analysis](#si).

### 2019-04-21 manuscript/06-supplement.Rmd
submitted revisions

# Supplemental Data {#si}

## *TIR1/AFB* genes {#AFBs} 
Auxin acts by binding to receptors (Auxin-signaling F-Boxes, or AFBs) that in turn target co-repressors (Aux/IAAs) for degradation. The six auxin receptor genes in the model plant *Arabidopsis thaliana*, *TIR1* and *AFB1-5*, evolved through gene duplication and diversification early in the history of vascular plants [@parry_complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@havens_synthetic_2012], and this rate sets the pace of lateral root development [@guseman_auxininduced_2015].

All members of this family have been shown to bind auxin and Aux/IAA proteins. However, AFB1 has drastically reduced ability to assemble into an SCF complex, due to the substitution E8K in its F-box domain, preventing it from inducing degradation of Aux/IAAs [@yu_untethering_2015]. This lack of SCF formation may allow for the high and ubiquitous AFB1 accumulation observed in *Arabidopsis* tissues [@parry_complex_2009]. Higher order receptor mutants in the family containing *afb1* mutants suggest that *AFB1* has a moderate positive effect on auxin signaling [@dharmasiri_plant_2005]. Additionally, AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while other family members preferentially bind indole-3-acetic acid [@prigge_arabidopsis_2016]. Interestingly, the strength and rate with which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable [@calderonvillalobos_combinatorial_2012; @havens_synthetic_2012]. AFB2 induces the degradation of certain Aux/IAA proteins at a faster rate than TIR1, suggesting some functional specificity has arisen since the initial duplication between the *TIR1/AFB1* and *AFB2/AFB3* clades.

@@

Examining the natural sequence variation across the *AFB* family \@ref(fig:AFBtree) revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity, hinting at their likely functional importance and bringing in to question the inconclusive role of *AFB1* in auxin signaling. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sister genes, *AFB2* and *AFB5* were more conserved. This matches our current understanding of *AFB3* as playing a minor role in the auxin signaling pathway [@dharmasiri_plant_2005] and suggests *AFB4* may be undergoing pseudogenization, especially when paired with its low expression levels [@prigge_arabidopsis_2016].

@@

Although most known functional regions are highly conserved in *AFB1*, there are some nonsynonymous polymorphism in the oligomerization domain that are only present in single accessions (F125E in Can-0 and I163N in Pu2-23). Mutations in this domain of *TIR1* frequently have a semidominant effect on root phenotypes [@dezfulian_oligomerization_2016; @wright_insights_2017]. Characterization of this allele and accession may help determine the role of *AFB1* in this pathway.

```{r AFB1-map, fig.cap="(ref: AFB1-map)"}
knitr::include_graphics("app_figure/Map_figure.png")
```
(ref: AFB1-map) **Map of AFB1 Oligomerization domain variant accessions** (a) Map showing the two accessions with variants in the AFB1 oligomerization domain (b) The user selectable parameters of the mapping tab used to generate the map are provided as an example of using the SNP Mapping tab in the ViVa webtool. See Figure \@ref(fig:app-workflow) for details on filling in the parameters.

The AFB4 and AFB5 receptors have an N-terminal extension prior to their F-box domains. This extension had very high nonsynonymous diversity (Figure \@ref(fig:full-AFBs-alignment)), suggesting that this extension does not play an important functional role in these proteins. Additionally, two frameshift variants and one stop-gained variant were observed in *AFB4* supporting its pseudogenization.

## *TPL/TPR* genes {#TPLs}

The auxin signaling pathway utilizes the TOPLESS (TPL) and TOPLESS-related (TPR) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin [@szemenyei_topless_2008]. In *Arabidopsis thaliana* the five member *TPL/TPR* family includes *TPL* and *TPR1-4*. The resulting proteins are comprised of three structural domains: an N-terminal TPL domain and two WD-40 domains [@long_topless_2006]. TPL/TPR proteins are recruited to the AUX/IAA proteins through interaction with the conserved Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain [@szemenyei_topless_2008]. Canonical EAR domains have the amino acid sequence LxLxL, as found in most AUX/IAAs [@overvoorde_functional_2005]. TPL/TPR co-repressors bind EAR domains via their C-terminal to LisH (CTLH) domains found near their N-termini (citations of pre-structure founding papers/reviews). Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs [@martin-arevalillo_structure_2017; @ke_structural_2015]. The residues required for higher-order multimers of TPL tetramers have also been identified [@ma_d53_2017]. Additional interactions with transcriptional regulation and chromatin modifying machinery are likely mediated by two tandem beta propeller domains of TPL/TPRs.

@@

The high degree of conservation in the entire N-terminal domain underscores its importance in TPL/TPR function (Figure \@ref(fig:full-TPLs-alignment). For example, the initial *tpl-1* mutation (N176H) in the ninth helix is a dominant gain-of-function allele [@long_topless_2006], which is capable of binding wild-type TPL protein and inducing protein aggregation [@ma_d53_2017]. It is therefore understandable that this helix had very low diversity as nonsynonymous variants in this domain could act in a dominant negative fashion.

@@

# *ARF* genes {#ARFs}

@@

For all class A *ARF*s, the middle region of the protein was the predominant high diversity region (Figure \@ref(fig:full-ARFs-alignment)). In the analyzed natural variation, ARF7 had several expansions of polyglutamine sequences in the middle region. Polyglutamine regions are known to readily expand and contract throughout evolutionary time due to replication error, and variation in polyglutamine length can be acted on by natural selection and have phenotypic consequences [@press_overdue_2014]. The ARF DNA-binding domain had very few, low-diversity missense mutations, as did the critical residues of the PB1 domain. Considering the necessity of their conserved functions, the low level of variation in these key DNA and protein-protein interaction domains is expected.

@@

Many of the class B *ARF*s have very high ${\pi_N/\pi_S}$ ratios relative to the other *ARF*s. ARF23 has a truncated DNA-binding domain and had a high ${\pi_N/\pi_S}$ value of 4.1. ARF13 has many high-diversity nonsense variants and lacks a C-terminal PB1 domain. This high level of diversity, prevelance of high-frequency nonsense variants and frequent loss of critical domains, may suggest that several genes in this class are undergoing pseudogenization.

@@

Class C ARFs show low nucleotide diversity scores, with all ${\pi_N/\pi_S}$ values substantially lower than 1. *ARF16* was the most conserved, whereas its clade members (*ARF10*, *ARF17*) had scores at least four times higher. Structurally, all three members of Class C ARFs contain a canonical B3 DNA-binding domain, but only ARF10 and ARF16 contain a PB1 domain. The DNA binding domains exhibit overall low diversity. Of the PB1 domain containing class C ARFs, ARF16 exhibits several missense variants which are sporadically distributed, in contrast to the conserved PB1 domain of ARF10. This conservation in the PB1 domain of ARF10 and the DBD of ARF16 may suggest subfunctionalization in this family.

## *Aux/IAA* genes {#IAAs}

For simplicity we have included only the alignment of the class A Aux/IAAs in the main manuscript. For completion we include here the complete alignment of the family (Figure \@ref(fig:full-IAA-alignment)).

(ref:full-IAA-alignment) **Alignment of the complete Aux/IAA family.** Protein sequences were aligned with DECIPHER [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the allele of strongest effect at that position, with light colors having weaker effects on function and darker colors stronger effects. Red indicates missense variants. Color scale is explained in [Methods](#methods). In grayscale underneath the plot, alignment consensus is shown as measured by Evolutionary trace [@pluciennik_balcony_2018; @wilkins_evolutionary_2012], with high consensus positions in black and low consensus in white.

```{r IAA6-div-plot, fig.cap="(ref:IAA6-div-plot)"}
coding_variants <- VCF[VCF$Transcript_ID == gene_info[gene_info$tair_symbol == "IAA6", "transcript_ID"], ]
p <- plotCodingDiv(uniqueCodingVars = coding_variants)
p + annotate("rect", xmin = 106, xmax = 127, ymin = 0.001, ymax = 1, color = "lightgrey", alpha =.1) + annotate("text", x = 116.5, y = 0.5, label = "region of positive\nselection identified\nin Winkler et al.")
```
(ref:IAA6-div-plot) **IAA6 diversity plot.** Nucleotide diversity of variant positions throughout the IAA6 coding sequence are plotted and colored according to the effect of the variant alleles at each position. The region of positive selection identified by Winkler et al. is highlighted.

>>>>

# Supplemental Analysis {#si}

## Additional natural variation in the *Aux/IAA* genes {#IAAs}

For simplicity we have included only the alignment of the class A Aux/IAAs in the main manuscript. We include here the complete alignment of the family (Figure \@ref(fig:full-IAA-alignment)).

(ref:full-IAA-alignment) **Alignment of the complete Aux/IAA family.** Protein sequences were aligned with DECIPHER [@wright_decipher_2015] and variants were mapped to this alignment and colored according to the predicted functional effect of the allele of strongest effect at that position, with light colors having weaker effects on function and darker colors stronger effects. Red indicates missense variants. Color scale is further explained in [Methods](#methods). Alignment consensus is shown in grayscale underneath the plot as measured by Evolutionary trace [@pluciennik_balcony_2018; @wilkins_evolutionary_2012], with high consensus positions in black and low consensus in white.

```{r IAA6-div-plot, fig.cap="(ref:IAA6-div-plot)", fig.height=4}
coding_variants <- VCF[VCF$Transcript_ID == gene_info[gene_info$tair_symbol == "IAA6", "transcript_ID"], ]
p <- plotCodingDiv(uniqueCodingVars = coding_variants)
p + annotate("rect", xmin = 106, xmax = 127, ymin = 0.001, ymax = 1, 
             color = "lightgrey", alpha =.1) + 
  annotate("text", x = 116.5, y = 0.1, 
           label = 
             "region of positive selection\nidentified in Winkler et al.") + labs(color = "", y = "nucleotide diversity,\nlog scale") + theme(legend.position = "bottom", legend.text = element_text(size = 8), axis.title = element_text(size = 12)) + guides(color=guide_legend(nrow=5))
```
(ref:IAA6-div-plot) **IAA6 diversity plot.** Nucleotide diversity of variant positions throughout the IAA6 coding sequence are plotted and colored according to the effect of the variant alleles at each position. The region of positive selection identified by Winkler et al. is highlighted.

## Natural variation in the *TIR1/AFB* genes {#AFBs} 
Auxin acts by binding to receptors (Auxin-signaling F-Boxes, or AFBs) that in turn target co-repressors (Aux/IAAs) for degradation. The six auxin receptor genes in the model plant *Arabidopsis thaliana*, *TIR1* and *AFB1-5*, evolved through gene duplication and diversification early in the history of vascular plants [@parry_complex_2009]. The rate of co-repressor degradation is determined by the identity of both the receptor and co-repressor [@havens_synthetic_2012], and this rate sets the pace of lateral root development [@guseman_auxininduced_2015].

All members of this family have been shown to bind auxin and Aux/IAA proteins. However, AFB1 has a drastically reduced ability to assemble into an SCF complex, due to the substitution E8K in its F-box domain, preventing it from inducing degradation of Aux/IAAs [@yu_untethering_2015]. This lack of SCF formation may allow for the high and ubiquitous AFB1 accumulation observed in *Arabidopsis* tissues [@parry_complex_2009]. Higher order receptor mutants in the family containing *afb1* mutants suggest that *AFB1* has a moderate positive effect on auxin signaling [@dharmasiri_plant_2005]. Additionally, AFB4 and AFB5 have been shown to preferentially and functionally bind the synthetic auxin picloram, while other family members preferentially bind indole-3-acetic acid [@prigge_arabidopsis_2016]. Interestingly, the strength and rate with which TIR1/AFBs are able to bind and mark Aux/IAAs for degradation are variable [@calderonvillalobos_combinatorial_2012; @havens_synthetic_2012]. AFB2 induces the degradation of certain Aux/IAA proteins at a faster rate than TIR1, suggesting some functional specificity has arisen since the initial duplication between the *TIR1/AFB1* and *AFB2/AFB3* clades.

@@

Examining the natural sequence variation across the *AFB* family revealed that *TIR1* and *AFB1* both had very low nonsynonymous diversity (Figure \@ref(fig:AFBtree)), hinting at their likely functional importance and bringing in to question the inconclusive role of *AFB1* in auxin signaling. *AFB3* and *AFB4* had higher nonsynonymous diversity, while their sister genes, *AFB2* and *AFB5* were more conserved. This matches our current understanding of *AFB3* as playing a minor role in the auxin signaling pathway [@dharmasiri_plant_2005]. Two frameshift variants and one stop-gained (nonsense) variant were observed in *AFB4* supporting its pseudogenization., suggesting that *AFB4* may be undergoing pseudogenization, especially when paired with its low expression levels [@prigge_arabidopsis_2016]. AFB4 and AFB5 have an N-terminal extension prior to their F-box domains. This extension had very high nonsynonymous diversity (Figure \@ref(fig:full-AFBs-alignment)), suggesting that this extension does not play an important functional role in these proteins.

@@

Although most known functional regions are highly conserved in *AFB1*, there are some nonsynonymous polymorphism in the oligomerization domain that are only present in single accessions (F125E in Can-0 and I163N in Pu2-23, as shown in Figure \@ref(fig:AFB1-map)). Mutations in this domain of *TIR1* frequently have a semidominant effect on root phenotypes [@dezfulian_oligomerization_2016; @wright_insights_2017]. Characterization of this allele and accession may help determine the role of *AFB1* in this pathway.

```{r AFB1-map, fig.cap="(ref:AFB1-map)"}
knitr::include_graphics("app_figure/Map_figure.png")
```

(ref:AFB1-map) **Map of AFB1 Oligomerization domain variant accessions.** (a) Map showing the two accessions with variants in the AFB1 oligomerization domain (b) The user selectable parameters of the mapping tab used to generate the map are provided as an example of using the SNP Mapping tab in the ViVa webtool. See Figure \@ref(fig:app-workflow) for details on filling in the parameters.

## Natural variation in the *TPL/TPR* genes {#TPLs}

The auxin signaling pathway utilizes the TOPLESS (TPL) and TOPLESS-related (TPR) family of Gro/TLE/TUP1 type co-repressor proteins to maintain auxin responsive genes in a transcriptionally-repressed state in the absence of auxin [@szemenyei_topless_2008]. In *Arabidopsis thaliana* the five member *TPL/TPR* family includes *TPL* and *TPR1-4*. The resulting proteins are comprised of three structural domains: an N-terminal TPL domain and two WD-40 domains [@long_topless_2006]. TPL/TPR proteins are recruited to the AUX/IAA proteins through interaction with the conserved Ethylene-responsive element binding factor-associated amphiphilic repression (EAR) domain [@szemenyei_topless_2008]. Canonical EAR domains have the amino acid sequence LxLxL, as found in most AUX/IAAs [@overvoorde_functional_2005]. TPL/TPR co-repressors bind EAR domains via their C-terminal to LisH (CTLH) domains found near their N-termini [@long_topless_2006]. Recent structural analyses of the TPL N-terminal domain have highlighted the precise interaction interface between TPL and AUX/IAA EAR domains, as well as the TPL-TPL dimerization and tetramerization motifs [@martin-arevalillo_structure_2017; @ke_structural_2015]. The residues required for higher-order multimers of TPL tetramers have also been identified [@ma_d53_2017]. Additional interactions with transcriptional regulation and chromatin modifying machinery are likely mediated by two tandem beta propeller domains of TPL/TPRs.

@@

The high degree of conservation in the entire N-terminal domain underscores its importance in TPL/TPR function (Figure \@ref(fig:full-TPLs-alignment)). For example, the initial *tpl-1* mutation (N176H) in the ninth helix is a dominant gain-of-function allele [@long_topless_2006], which is capable of binding wild-type TPL protein and inducing protein aggregation [@ma_d53_2017]. It is therefore understandable that this helix had very low diversity as nonsynonymous variants in this domain could act in a dominant negative fashion.

@@

## Natural Variation in the *ARF* genes {#ARFs}

@@

For all class A *ARF*s, the middle region of the protein was the predominant high diversity region (Figure \@ref(fig:full-ARFs-alignment)). In the analyzed natural variation, ARF7 had several expansions of polyglutamine sequences in the middle region. Polyglutamine regions are known to readily expand and contract throughout evolutionary time due to replication error, and variation in polyglutamine length can have phenotypic consequences and be acted on by natural selection [@press_overdue_2014]. The ARF DNA-binding domain had very few, low-diversity missense mutations, as did the critical residues of the PB1 domain. Considering the necessity of their conserved functions, the low level of variation in these key DNA and protein-protein interaction domains is expected.

@@

Many of the class B *ARF*s have very high ${\pi_N/\pi_S}$ ratios relative to the other *ARF*s. ARF23 has a truncated DNA-binding domain and had a high ${\pi_N/\pi_S}$ value of 4.1. ARF13 has many high-diversity nonsense variants and lacks a C-terminal PB1 domain. This high level of diversity, prevalance of high-frequency nonsense variants and frequent loss of critical domains, may suggest that several genes in this class are undergoing pseudogenization.

@@

Class C ARFs show low nucleotide diversity scores, with all ${\pi_N/\pi_S}$ values substantially lower than 1. *ARF16* was the most conserved, whereas the other clade members (*ARF10*, *ARF17*) had scores at least four times higher. Structurally, all three members of Class C ARFs contain a canonical B3 DNA-binding domain, but only ARF10 and ARF16 contain a PB1 domain. The DNA binding domains exhibited overall low diversity. Of the PB1 domain containing class C ARFs, ARF16's PB1 domain exhibited several missense variants which are sporadically distributed, in contrast to the conserved PB1 domain of ARF10. This conservation in the PB1 domain of ARF10 and the DBD of ARF16 may suggest subfunctionalization in this class.

### 2019-04-21 manuscript/index.Rmd
submitted revisions

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE,
  collapse = TRUE, 
  comment = "#>", 
  tidy = TRUE, 
  cache = TRUE, 
  cache.lazy = FALSE, 
  fig.height=4, 
  fig.width=7, 
  dpi = 600, 
  out.extra='style="margin: auto; display: block; padding-top: 15px;"')

>>>>

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE,
  collapse = TRUE, 
  comment = "#>", 
  tidy = TRUE, 
  cache = TRUE, 
  cache.lazy = FALSE, 
  fig.height=4, 
  fig.width=7, 
  dpi = 800, 
  out.extra='style="margin: auto; display: block; padding-top: 15px;"')
